\section{Hospital Markov chain model}
The following Markov chain represents the transition between states of a hospital while capturing the EMS interaction with it. The hospital accepts both ambulance and other patients normally until a certain threshold \(T\) is reached. When it is reached all ambulances that arrive will be marked as \textit{"parked outside"} until the number of people in the system is reduced below \(T\). Additionally, if the patients in the hospital keep rising, they may exceed the number of servers \(C\) available, which will in turn mean that every new patient will have to wait for a server to become free. The states of the Markov chain are denoted by \((u,v)\) where:

\begin{itemize}
    \item \(u\) = number of ambulances parked outside of the hospital
    \item \(v\) = number of patients in the hospital
\end{itemize}


\begin{figure}
    \centering
    \begin{tikzpicture}[-, node distance = 0.9cm, auto, every node/.style={scale=0.7}]

        % Markov chain variables
        \tikzmath{
            let \initdist = 0.5cm;
            let \altdist = 1.2cm;
            let \minsz = 1.6cm;
        }

        % S_1 and S_2 rectangles
        \tikzmath{
            let \leftOne = -0.8;
            let \rightOne = 2.7;
            let \upOne = 0.8;
            let \downOne = -2.7;
            let \leftTwo = 2.8;
            let \rightTwo = 13;
            let \upTwo = -2.95;
            let \downTwo = -16.4;
        }

        % General case variables
        \tikzmath{
            let \GCsmallx = 8.3;
            let \GCsmally = -9.5;
            let \GCbigx = 4.1;
            let \GCbigy = -11.8;
        }

        % Rectangle for S1
        \draw[ultra thin, dashed] (\leftOne, \downOne) -- (\leftOne, \upOne);
        \draw[ultra thin, dashed] (\leftOne, \upOne) -- (\rightOne, \upOne);
        \draw[ultra thin, dashed] (\rightOne, \upOne) -- node {\Huge{\( \quad S_1 \)}}(\rightOne, \downOne);
        \draw[ultra thin, dashed] (\rightOne, \downOne) -- (\leftOne, \downOne);

        % Rectangle for S2
        \draw[ultra thin, dashed] (\leftTwo, \downTwo) -- node {\Huge{\( S_2 \quad \)}}(\leftTwo, \upTwo);
        \draw[ultra thin, dashed] (\leftTwo, \upTwo) -- (\rightTwo, \upTwo);
        \draw[ultra thin, dashed] (\rightTwo, \upTwo) -- (\rightTwo, \downTwo);
        \draw[ultra thin, dashed] (\rightTwo, \downTwo) -- (\leftTwo, \downTwo);

        % Small square of general case
        \draw [thick] (\GCsmallx, \GCsmally) -- node {} (\GCsmallx + 0.4, \GCsmally);
        \draw [thick] (\GCsmallx + 0.4, \GCsmally) -- node {} (\GCsmallx + 0.4, \GCsmally - 0.4);
        \draw [thick] (\GCsmallx + 0.4, \GCsmally - 0.4) -- node {} (\GCsmallx, \GCsmally - 0.4);
        \draw [thick] (\GCsmallx, \GCsmally - 0.4) -- node {} (\GCsmallx, \GCsmally);


        % Dashed lines to from small square to big one 
        \draw [ultra thin] (\GCsmallx, \GCsmally) -- node {} (\GCbigx, \GCbigy);
        \draw [ultra thin] (\GCsmallx + 0.4, \GCsmally) -- node {} (\GCbigx + 4, \GCbigy);
        \draw [ultra thin] (\GCsmallx, \GCsmally - 0.4) -- node {} (7, \GCbigy);
        \draw [ultra thin] (\GCsmallx + 0.4, \GCsmally - 0.4) -- node {} (\GCbigx + 4, \GCbigy - 4);
        
        % Big Square of general case
        \draw [ultra thick] (\GCbigx, \GCbigy) -- node {} (\GCbigx + 4, \GCbigy);
        \draw [ultra thick] (\GCbigx + 4, \GCbigy) -- node {} (\GCbigx + 4, \GCbigy - 4);
        \draw [ultra thick] (\GCbigx + 4, \GCbigy - 4) -- node {General Case} (\GCbigx, \GCbigy - 4);
        \draw [ultra thick] (\GCbigx, \GCbigy - 4) -- node {} (\GCbigx, \GCbigy);

        % First Line
        \node[state, minimum size=1.5cm] (zero) {(0,0)};
        \node[state, node distance = \initdist, minimum size=\minsz, below right=of zero] (one) {(0,1)};
        \node[draw=none, node distance = \initdist, minimum size=\minsz, below right=of one] (two) {\textbf{\( \ddots \)}};
        \node[state, node distance = \initdist, minimum size=\minsz, below right=of two] (three) {(0,T)};
        \node[state, node distance = \altdist, minimum size=\minsz, right=of three] (four) {(0,T+1)};
        \node[draw=none, node distance = \altdist, minimum size=\minsz, right=of four] (five) {\textbf{\dots}};
        \node[state, minimum size=\minsz, right=of five] (six) {(0,C)};
        \node[draw=none, minimum size=\minsz, right=of six] (seven) {\textbf{\dots}};

        % Second Line
        \node[state, minimum size=\minsz, below=of three] (three_one) {(1,T)};
        \node[state, minimum size=\minsz, below=of four] (four_one) {(1,T+1)};
        \node[draw=none, minimum size=\minsz, below=of five] (five_one) {\textbf{\dots}};
        \node[state, minimum size=\minsz, right=of five_one] (six_one) {(1,C)};
        \node[draw=none, minimum size=\minsz, right=of six_one] (seven_one) {\textbf{\dots}};
        
        % Third Line
        \node[state, minimum size=\minsz, below=of three_one] (three_two) {(2,T)};
        \node[state, minimum size=\minsz, below=of four_one] (four_two) {(2,T+1)};
        \node[draw=none, minimum size=\minsz, below=of five_one] (five_two) {\textbf{\dots}};
        \node[state, minimum size=\minsz, right=of five_two] (six_two) {(2,C)};
        \node[draw=none, minimum size=\minsz, right=of six_two] (seven_two) {\textbf{\dots}};

        % Fourth line
        \node[draw=none, node distance = \altdist, minimum size=\minsz, below=of three_two] (three_three) {\textbf{\vdots}};
        \node[draw=none, node distance = \altdist, minimum size=\minsz, below=of four_two] (four_three) {\textbf{\vdots}};
        \node[draw=none, node distance = 2cm, minimum size=\minsz, below=of five_two] (five_three) {};
        \node[draw=none, node distance = \altdist, minimum size=\minsz, below=of six_two] (six_three) {\textbf{\vdots}};

        % Fifth line
        % \node[state, node distance = \altdist, minimum size=\minsz, below=of five_three] (general_case_mid) {\( (u_i, v_i) \)};
        \node[draw=none, node distance = 0.3cm, minimum size=\minsz, below=of four_three] (general_case_up) {};
        \node[state, node distance = \altdist, minimum size=\minsz, below=of general_case_up] (general_case_mid) {\( (u_i, v_i) \)};

        \node[draw=none, node distance = \altdist, minimum size=\minsz, below=of general_case_mid] (general_case_down) {};
        \node[draw=none, node distance = \altdist, minimum size=\minsz, left=of general_case_mid] (general_case_left) {};
        \node[draw=none, node distance = \altdist, minimum size=\minsz, right=of general_case_mid] (general_case_right) {};

        \draw[every loop]
            % First Horizontal Edges
            (zero) edge[bend left] node {\( \Lambda \)} (one)
            (one) edge[bend left] node {\( \mu \)} (zero)
            (one) edge[bend left] node {\( \Lambda \)} (two)
            (two) edge[bend left] node {\( 2 \mu \)} (one)
            (two) edge[bend left] node {\( \Lambda \)} (three)
            (three) edge[bend left] node {\( T \mu \)} (two)
            (three) edge[bend left] node {\( \lambda^o \)} (four)
            (four) edge[bend left] node {\( (T+1) \mu \)} (three)
            (four) edge[bend left] node {\( \lambda^o \)} (five)
            (five) edge[bend left] node {\( (T+2) \mu \)} (four)
            (five) edge[bend left] node {\( \lambda^o \)} (six)
            (six) edge[bend left] node {\( C\mu \)} (five)
            (six) edge[bend left] node {\( \lambda^o \)} (seven)
            (seven) edge[bend left] node {\( C\mu \)} (six)

            % Second Horizontal Edges
            (three_one) edge[bend left] node {\( \lambda^o \)} (four_one)
            (four_one) edge[bend left] node {\( (T+1) \mu \)} (three_one)
            (four_one) edge[bend left] node {\( \lambda^o \)} (five_one)
            (five_one) edge[bend left] node {\( (T+2) \mu \)} (four_one)
            (five_one) edge[bend left] node {\( \lambda^o \)} (six_one)
            (six_one) edge[bend left] node {\( C\mu \)} (five_one)
            (six_one) edge[bend left] node {\( \lambda^o \)} (seven_one)
            (seven_one) edge[bend left] node {\( C\mu \)} (six_one)

            % Third Horizontal Edges
            (three_two) edge[bend left] node {\( \lambda^o \)} (four_two)
            (four_two) edge[bend left] node [below] {\( (T+1) \mu \)} (three_two)
            (four_two) edge[bend left] node {\( \lambda^o \)} (five_two)
            (five_two) edge[bend left] node {\( (T+2) \mu \)} (four_two)
            (five_two) edge[bend left] node {\( \lambda^o \)} (six_two)
            (six_two) edge[bend left] node {\( C\mu \)} (five_two)
            (six_two) edge[bend left] node {\( \lambda^o \)} (seven_two)
            (seven_two) edge[bend left] node {\( C\mu \)} (six_two)

            % First Vertical Edges
            (three) edge[bend left] node {\( \lambda^A \)} (three_one)
            (three_one) edge[bend left] node {\( T \mu \)} (three)
            (three_one) edge[bend left] node {\( \lambda^A \)} (three_two)
            (three_two) edge[bend left] node {\( T\mu \)} (three_one)
            (three_two) edge[bend left] node {\( \lambda^A \)} (three_three)
            (three_three) edge[bend left] node {\( T\mu \)} (three_two)

            % Second Vertical Edges
            (four) edge node {\( \lambda^A \)} (four_one)
            (four_one) edge node {\( \lambda^A \)} (four_two)
            (four_two) edge node {\( \lambda^A \)} (four_three)

            % Fourth Vertical Edges
            (six) edge node {\( \lambda^A \)} (six_one)
            (six_one) edge node {\( \lambda^A \)} (six_two)
            (six_two) edge node {\( \lambda^A \)} (six_three)

            % General Case
            (general_case_left) edge[bend left] node {\( \lambda^o \)} (general_case_mid)
            (general_case_mid) edge[bend left] node {\( v_i \mu \)} (general_case_left)
            (general_case_right) edge[bend left] node {\( (v_i +1) \mu \)} (general_case_mid)
            (general_case_mid) edge[bend left] node {\( \lambda_o \)} (general_case_right)
            % (five_three) edge node {\( \lambda_A \)} (general_case_mid)
            (general_case_up) edge node {\( \lambda_A \)} (general_case_mid)
            (general_case_mid) edge node {\( \lambda_A \)} (general_case_down)
            ;
    \end{tikzpicture}
    \caption{Markov chain} 
    \label{markov_model}
\end{figure}


\subsection{Markov-chain state mapping function}
The transition matrix of the Markov-chain representation described above can be denoted by a state mapping function. The state space of this function is defined as:



\begin{align}
    S(T) =& S_1(T) \cup S_2(T) \text{ where:} \nonumber \\
    S_1(T) =& \left\{(0, v)\in\mathbb{N}_0^2 \; | \; v < T \right\} \\
    S_2(T) =& \{(u, v)\in\mathbb{N}_0^2 \; | \; v \geq T \} \nonumber
\end{align}

Therefore, the entries of the transition matrix \(Q\), can be given by \( q_{i,j} = q_{(u_i, v_i),(u_j, v_j)} \) which is the transition rate from state \( i = (u_i, v_i) \) to state \( j = (u_j , v_j) \) for all \( (u_i, v_i), (u_j, v_j) \in S \).

\begin{equation}
    q_{i, j} = 
    \begin{cases}
        \Lambda, & \textbf{if } (u_i, v_i) - (u_j, v_j) = (0,-1) \textbf{ and } v_i < \text{t} \\
        \lambda^o, & \textbf{if } (u_i, v_i) - (u_j, v_j) = (0,-1) \textbf{ and } v_i \geq \text{t} \\
        \lambda^a, & \textbf{if } (u_i, v_i) - (u_j, v_j) = (-1,0) \\
        v_i \mu, & \textbf{if } (u_i, v_i) - (u_j, v_j) = (0,1) \textbf{ and } v_i \leq C \textbf{ or} \\ & \hspace{0.37cm}(u_i, v_i) - (u_j, v_j) = (1,0) \textbf{ and } v_i = T \leq C \\
        C \mu, & \textbf{if } (u_i, v_i) - (u_j, v_j) = (0,1) \textbf{ and } v_i > C \textbf{ or} \\ & \hspace{0.37cm}(u_i, v_i) - (u_j, v_j) = (1,0) \textbf{ and } v_i = T > C\\
        % T \mu, & \textbf{if } (u_i, v_i) - (u_j, v_j) = (1,0) \textbf{ and } v_i = \text{t} \\
        -\sum_{j=1}^{|Q|}{q_{i,j}} & \textbf{if } i = j \\
        0, & \textbf{otherwise}
    \end{cases}
\end{equation}

In order to acquire an exact solution of the problem a slight adjustment needs to be considered. The problem defined above assumes no upper boundary to the number of people that can wait for service or the number of ambulances that can be parked outside. Therefore, a different state space \( \tilde S \) needs to be constructed where \( \tilde S \subseteq S \) and there is a maximum allowed number of people \( N \) that can be in the system and a maximum allowed number of ambulances \( M \) parked outside:

\begin{equation}
    \tilde S = \left\{ (u, v) \in S\;| u \leq M, v\leq N \right\}
\end{equation}


\subsection{Steady State}
Having calculated the transition matrix \( Q \) for a given set of parameters the probability vector \( \pi \) needs to be considered. The vector \( \pi \) is commonly used to study such stochastic systems and it's main purpose is to keep track of the probability of being at any given state of the system. The term \textit{steady state} refers to the instance of the vector \( \pi \) where the probabilities of being at any state become stable over time. Thus, by considering the steady state vector \( \pi \) the relationship between it and \(Q \) is given by:

\[
\frac{d\pi}{dt} = \pi Q = 0
\]

There are numerous methods that can be used to solve problems of such kind. In this paper only numeric and algebraic approaches will be considered. 

\subsubsection{Numeric integration}
The first approach to be considered is to solve the differential equation numerically by observing the behaviour of the model over time. The solution is obtained via python's SciPy library. The functions odeint and solve\textunderscore ivp have been used in order to find a solution to the problem. Both of these functions can be used to solve any system of first order ODEs.

\subsubsection{Linear algebraic approach}
Another approach to be considered is the linear algebraic method. The steady state vector can be found algebraically by satisfying the following set of equations:
\[ \pi Q = 0 \]
\[ \sum_{i} \pi_i = 1 \]

These equations can be solved by slightly altering \( Q \) such that the final column is replaced by a vector of ones. Thus, the resultant solution occurs from solving the equation \( \tilde{Q}^T \pi = b \) where \( \tilde{Q} \) and \( b \) are defined as:

\[
\tilde{q_{i, j}} = 
\begin{cases}
    1, & \textbf{if } j = |Q| \\
    q_{i,j}, & \textbf{otherwise}
\end{cases}
\]

\[
b = 
\begin{bmatrix}
    x_{1} \\
    x_{2} \\
    \vdots \\
    x_{m}
\end{bmatrix}
\]


\subsubsection{Least Squares approach}
Finally, the last approach to be considered is the least squares method. This approach is considered because while the problem becomes more complex (in terms of input parameters) the computational time required to solve it increases exponentially. Thus, one may obtain the steady state vector \( \pi \) by solving the following equation.

\[
\pi = \text{argmin}_{x\in\mathbb{R}^{d}}\|Mx-b\|_2^2
\]

\newpage
\subsection{Expressions derived from \( \pi \):}
One may easily derive the average number of individuals that are at any given state using \( pi \). The average number of individuals in state \( i \) can be calculated by multiplying the number of individuals that are present in state \( i \) with the probability of being at that particular state (i.e \(\pi_i (u_i + v_i)\)). Using this logic it is possible to calculate any performance measures that are related to the mean number of individuals in the system.


Average number of patients in the system: 
\begin{equation}
    L = \sum_{i=1}^{|\pi|} \pi_i (u_i + v_i)
\end{equation} 

Average number of patients in the hospital: 
\begin{equation}
    L_H = \sum_{i=1}^{|\pi|} \pi_i v_i
\end{equation}

Average number of ambulances being blocked:
\begin{equation}
    L_A = \sum_{i=1}^{|\pi|} \pi_i u_i
\end{equation}

Consequently getting the performance measures that are related to the duration of time is not as straightforward. Such performance measures are the mean waiting time in the system and the mean time blocked in the system. Under the scope of this study two approaches have are considered; a recursive algorithm and consequently a closed-form formula.

\subsubsection{Mean Waiting Time - Recursive}
The research question that needs to be answered here is: ``When an ambulance/other patient enters the system, what is the expected time that they will have to wait?''. In order to formulate the answer to that question one needs to consider all possible scenarios of what state the system can be in when an individual arrives. 

To calculate such times one must first identify all states \((u_i, v_i)\) that when the model is in that state and an individual arrives they will incur a wait. For this particular Markov chain, this points to all states that satisfy \( v_i \geq C \) i.e. all states where the number of individuals in the hospital exceed the number of servers. It is important to note here that an this is not a strict inequality since when the model is at the state with $C$ patients, the arriving patient will proceed to wait. The set of such states is defined as \textit{waiting states} and can be denoted as a subset of all the states, where:

\[
S_w = \{(u, v) \in S \; | \; v > C \}    
\]

Additionally, there are certain states in the model where arrivals cannot occur. An "ambulance" patient cannot arrive in the model when the model is at any state $(M, v) \; \forall \; v \geq T$ and equivalently an "other" patient cannot arrive whenever the model is at any state $(u, N) \; \forall \; u$ where $N$ is the system capacity and $M$ is the parking capacity. In essence this indicates that the formula will behave differently for ambulance and other patients. Therefore the set of all such states that an arrival may occur are defined as \textit{accepting states} and are denoted as:

\[
S_A^{(a)} = \{(u, v) \in S \; | \; v < M \}
\]

\[
S_A^{(o)} = \{(u, v) \in S \; | \; u < N \}
\]

Moreover, another element that needs to be considered is the expected waiting time in each state $c(u,v)$. In order to do so a variation of the Markov model has to be considered where when the individual arrives at any of the states of the model no more arrivals can occur after that. Thus, one may acquire the desired time by calculating the inverse of the sum of the out-flow rate of that state. Since we are ignoring arrivals though the only way to exit the state will only be via a service. In essence this notion can be expressed as:

\begin{equation} \label{eq:ExpectedWaitState}
    c(u,v) = \frac{1}{\text{min}(v,C)\mu}    
\end{equation}

Equation \ref{eq:ExpectedWaitState} considers two scenarios of states; whether the number of patients are more than the number of servers or not. Now, using the above equations, and considering all sates that belong in \(S_w\) the following recursive formula can be used to get the mean waiting time spent in each state in the Markov model. Note that the formula is different for each type of patients. Thus, the expected waiting time of an "other" patient when they arrive at state \((u,v)\) is given by:

\begin{equation}
    w^{(o)}(u,v) = 
    \begin{cases} 
        0, \hspace{4.85cm} & \textbf{if } (u,v) \notin S_w \\
        c(u,v + 1) + w(0, v-1), & \textbf{otherwise} \\
    \end{cases}
\end{equation}

The equivalent time for "ambulance" patients is denoted in a slightly different manner. States \(s_a\) and \(s_n\) represent the \textit{arriving state} and the \textit{next state} to be visited by the recursive formula. The arriving state is the state that the individual will actually arrive in when the model is in state \((u,v)\) and the next state is the one that the patient will move down to. Note here that all \textit{next states} move straight to the first row since no wait occurs for any ambulance patient in any row apart from the first one.
\begin{equation}
    w^{(a)}(u,v) = 
    \begin{cases}
        0 & \textbf{if } (u,v) \notin S_w \\
        c(s_a) + w(s_n) & \textbf{otherwise}
    \end{cases}
\end{equation}

\begin{multicols}{2}
    \begin{equation*}
        s_a = 
        \begin{cases} 
            (u + 1, v), & \textbf{if } v \geq T \\
            (u, v + 1), & \textbf{otherwise}
        \end{cases}
    \end{equation*}

    \begin{equation*}
        s_n = 
        \begin{cases} 
            (0, T), & \textbf{if } u \geq 1 \\
            (0, v - 1), & \textbf{otherwise}
        \end{cases}
    \end{equation*}
\end{multicols}


Thus, the overall mean waiting time can be calculated by summing over all expected waiting times of accepting states multiplied by the probability of being at that state and dividing by the probability of an individual arriving in the system. This formula has to be performed twice, one for each type of patient:

\begin{equation}
    W^{(o)} = \frac{\sum_{(u,v) \in S_A^{(o)}} w^{(o)}(u,v) \pi_{(u,v)}}{\sum_{(u,v) \in S_A^{(o)}} \pi_{(u,v)}}
\end{equation}

\begin{equation}
    W^{(a)} = \frac{\sum_{(u,v) \in S_A^{(a)}} w^{(a)}(u,v) \pi_{(u,v)}}{\sum_{(u,v) \in S_A^{(a)}} \pi_{(u,v)}}
\end{equation}



% \begin{equation}
%     w(u,v) = 
%     \begin{cases}
%         w_E(u,v) + w(u-1,v),  & \textbf{if } (u,v) \in S_w \textbf{ and } v = T \textbf{ and } u > 0\\
%         w_E(u,v) + w(u,v-1), & \textbf{if } (u,v) \in S_w \textbf{ and } v \neq T \textbf{ or } u = 0\\
%         0, & \textbf{otherwise}
%     \end{cases}
% \end{equation}

Thus, the overall mean waiting time can be calculated by summing all arrivals that will have a wait and dividing by the probability of an individual having to wait:

\begin{equation}
    W = \sum_{(u,v) \in S} w(u,v) \pi_{(u,v)}
\end{equation}

\subsubsection{Mean Waiting Time - Closed-form}














\newpage
Mean waiting time in the hospital:
\begin{equation}
    W_q = \sum_{i=1}^{|\pi|} \frac{\text{max}(v_i - C, 0) \; \pi_i}{\sum_{\substack{j=1 \\ i \neq j}}^{|\pi|} q_{i j}}
\end{equation}

\begin{equation}
    W_q = \sum_{i=1}^{|\pi|} \pi_i \frac{v_i - c}{v_i \mu}, \quad v_i > c
\end{equation}
